Mathematical description of Naomi

### 2022-12-06 src/example_naomi/maths.Rmd
Wow, more maths!

# Notation

* District $x$
  * Located in Spectrum region file $R_x$
* Sex $s \in \{\text{F}, \text{M}\}$
* Age $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$
* Time $t \in \{T_1, T_2, T_3\}$
  * $T_1$: Most recent national household survey with HIV testing
  * $T_2$: Current time period at which to generate estimates
  * $T_3$: Short-term project period (typically 9-12 months)
  * Note $T_{1:3}$ specified to nearest calendar quarter
* Population size $N_{x, s, a, t}$
* HIV prevalence $\rho_{x, s, a, t}$
* ART coverage $\alpha_{x, s, a, t}$
* Annual HIV incidence rate $\lambda_{x, s, a, t}$

# HIV prevalence and ART coverage at $T_1$

## Background

* $u \sim \text{ICAR}(\sigma)$ refers to the intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
* $u \sim \text{BYM}2(\sigma, \phi)$ refers to a reparameterised Besag-York-Mollie model (BYM2) with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
* For both the ICAR and BYM2 models, we follow recommendations on scaling, disconnected adjacency graph components, and islands.

## HIV prevalence

@@

## ART coverage

@@

## HIV incidence rate

@@

* ART coverage is modelled by
$$
\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha+\beta_{T_2, s=M}^\alpha+u_{x, T 2}^\alpha+u_{x, a<15, T_2}^\alpha+\eta_{R_x, s, a, T 2}^\alpha
$$

## Clarification notes to discuss with Jeff

* Some confusion on how to offset terms work -- they seem cool though. Might prefer they not be called $\eta$ which is in my head linear predictor
  * $\eta^\lambda_{R_x, s, a, t}$ aren't these $R_x, s, a, t$ specific rather than 
* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49
* Sometimes in subscript we have $x, s, a, t$ and sometimes e.g. $x, s, at$ -- why?
* $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
* Inconsistent use of $T2$ and $T_2$

>>>>

# Process specification

## Notation

* District $x$
  * Located in Spectrum region file $R_x$
* Sex $s \in \{\text{F}, \text{M}\}$
* Age $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$
* Time $t \in \{T_1, T_2, T_3\}$
  * $T_1$: Most recent national household survey with HIV testing
  * $T_2$: Current time period at which to generate estimates
  * $T_3$: Short-term project period (typically 9-12 months)
  * Note $T_{1:3}$ specified to nearest calendar quarter
* Population size $N_{x, s, a, t}$
* HIV prevalence $\rho_{x, s, a, t}$
* ART coverage $\alpha_{x, s, a, t}$
* Annual HIV incidence rate $\lambda_{x, s, a, t}$
* $u \sim \text{ICAR}(\sigma)$ refers to the intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$.
* $u \sim \text{BYM}2(\sigma, \phi)$ refers to a reparameterised Besag-York-Mollie model (BYM2) with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
* For both the ICAR and BYM2 models, we follow recommendations on scaling, disconnected adjacency graph components, and islands.

## HIV prevalence and ART coverage at $T_1$

@@

@@

### HIV incidence rate

@@

Change in ART coverage is modelled by
$$
\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \beta_{T_2}^\alpha + \beta_{T_2, s = \text{M}}^\alpha + u_{x, {T_2}}^\alpha + u_{x, a < 15, {T_2}}^\alpha + \eta_{R_x, s, a, {T_2}}^\alpha
$$
where

* $\eta_{R_x, s, a, {T_2}}^\alpha$ are offsets for the logit change in ART coverage between $T_1$ and $T_2$ from Spectrum region $R_x$
* $\beta_{T_2}^\alpha$ is the average change in logit ART coverage between $T_1$ and $T_2$ for adult women
* $\beta_{T_2, s = \text{M}}^\alpha$ is the difference in average change for men compared to women
* $u_{x, {T_2}}^\alpha \sim \mathcal{N}(0, \sigma_{XT}^\alpha)$ are district random effects
* $u_{x, a < 15, {T_2}}^\alpha$ is the difference in logit ART coverage for children compared to adults

**Flag for not understanding** ART coverage at $T_3$ is modelled by 
$$
\text{logit}(\alpha_{x, s, a, {T_2}}) = \text{logit}(\alpha_{x, s, a, {T_1}}) + \eta_{R_x, s, a, {T_3}}^\alpha
$$

## ANC testing cascade

* For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
$$
\log (\Psi_{x, a, t}) = \log (N_{x, \text{F}, a, t}) + \psi_{R_x, a, t} + \beta^\psi + u_x^\psi
$$
  * $N_{x, \text{F}, a, t}$ are the female population sizes
  * $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$
  * $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility
  * $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects
* HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
\begin{align}
\text{logit}(\rho_{x, a, t}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a, t}) + \beta^{\rho^{\text{ANC}}} + \beta_{t \in(T_2, T_3\}}^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + u_{x, t \in\{T_2, T_3\}}^{\rho^{\text{ANC}}} + \eta_{R_x, a, t}^{\rho^{\text{ANC}}} \\
\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}}
\end{align}
  * $\eta_{R_x, a, t}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$
  * $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes at $T_1$ after removing the offset
  * $\beta_{t \in(T_2, T_3\}}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the change in average difference from $T_1$ to $T_2$
  * $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects
  * $u_{x, t \in\{T_2, T_3\}}^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_{XT}^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ capture the change in district random effects between times $T_1$ and $T_2$

## ANC attendance

* Number of PLHIV on ART is $A_{x, s, a, t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}$
* Let $\gamma_{x, x', t} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$ at time $t$
* We assume that $\gamma_{x, x', t} = 0$ for $x \notin \{x, \text{ne}(x)\}$ i.e. that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$ where $\sim$ is an adjacency relation
  * $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x', t} = 1$
* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is the log odds ratio of seeking ART in the district $x'$ neighbouring $x$ **flag for confusion here**
\begin{align}
\tilde \gamma_{x, t} &= \tilde \gamma_0 + u_x^{\tilde \gamma} + u_x^{\tilde \gamma, t = T_2} \\ 
\tilde \gamma_0 &= -4 \\
u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma}) \\
u_x^{\tilde \gamma, t \in \{T_2, T_3\}} &\sim \mathcal{N}(0, \sigma_{XT}^{\tilde \gamma}) 
\end{align}
  * $\tilde \gamma_{x, t}$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district
  * $u_x^{\tilde \gamma}$ are district random effects
  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is the change in district level log-odds ratios between $T_1$ and $T_2$ -- **flag for confusion: shouldn't this just be $t = T_2$ here**
  * Note: only one value estimated for each district, implying no preference between neighbouring districts, and no sex-age specific modelling
* Calculate multinomial probabilities via softmax (n.b. this is the sigmoid function, the inverse of the softmax)
$$
\gamma_{x, x', t} = \frac{1}{1 + \sum_{x^\star \sim x} \exp(\tilde \gamma_{x^\star, t})}
$$
* The number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a, t} = A_{x, s, a, t} \cdot \gamma_{x, x', t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t} \cdot \gamma_{x, x', t}$
* Total attending ART facilities in district $x'$ is $\tilde A_{x', s, a, t} = \sum_{x \sim x', x = x'} A_{x, x', s, a, t}$

## Awareness of HIV status

* The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a, t}$
* The proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a, t}$ such that
$$
v_{x, s, a, t} = \alpha_{x, s, a, t} + \frac{\tilde v_{R_x, s, a, t}}{1 - \alpha_{x, s, a, t}}
$$

# Likelihood specification

## Notation

* Use $\{x\}$ to denote a collection of districts, $\{s\}$ a collection of sexes and $\{a\}$ a collection of ages for which an outcome is observed

## Household survey data

* Let $\nu$ be a household survey occuring at $T_1$
* Furnishes weighted observations for HIV prevalence $\hat \rho_{\{x\}, \{s\}, \{a\}, \nu}$, ART coverage $\hat \alpha_{\{x\}, \{s\}, \{a\}, \nu}$ and proportion recently infected $\hat \kappa_{\{x\}, \{s\}, \{a\}, \nu}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$
* The observed cases are $Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$
* For HIV prevalence we use a binomial working likelihood
\begin{align}
Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t}}
\end{align}
* For ART coverage we also use a binomial working likelihood
\begin{align}
Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
\end{align}
* For recent infections we also use a binomial working likelihood
\begin{align}
Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\kappa_{x, s, a, t} &= 1 - \exp(- \lambda_{x, s, a, t} \cdot \frac{1 - \rho_{x, s, a, t}}{\rho_{x, s, a, t}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
\Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
\beta_T &\sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})
\end{align}

# Clarification notes to discuss with Jeff

* Not sure what $R_x$ means. Is it like $R_x: \mathcal{X} \to \mathcal{X}$
* Some confusion on how to offset terms work -- they seem cool though. Might prefer they not be called $\eta$ which is in my head linear predictor
  * $\eta^\lambda_{R_x, s, a, t}$ aren't these $R_x, s, a, t$ specific rather than 
* Aggregation over age 15-49 uses sum from $a = 15$ to $a = 45$? Perhaps this is because $a = 45$ actually corresponds to the age band 45-49. Edit: yes, $a = 45$ corresponds to 45-49
* Sometimes in subscript we have $x, s, a, t$ and sometimes e.g. $x, s, at$ -- why?
* Why $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ rather than e.g. $\sigma^\lambda \sim \mathcal{N}^+(0, 2.5)$
* Inconsistent use of $T2$ and $T_2$ in the appendix (I prefer the subscript so try to use throughout here)
* In softmax function, confused as to what this equation is saying: shouldn't there be an equals sign in there before the large brackets?
